---
title: "EC425_P1"
author: "Nathan Bracken"
date: "4/14/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(here, stargazer, foreign, sandwich)
```

# Description  
In this empirical project, you will analyze experimental data from the Oregon Health Plan Experiment (OHP) which we discussed in class. In particular, you will analyze real data from in- person interviews conducted in the Portland, Oregon, metropolitan area. The interviews were conducted about 25 months after the OHP lottery. The interview included detailed questionnaires on insurance coverage, health care use, health status, inventory of medications. The interview also contains questions that help assess depression and health-related quality of life. Finally, performance of anthropometric and blood-pressure measurements is taken, and dried blood spots are obtained. For more information about OHP and the in-person interviews, see here and here. 

# Instructions  
Please submit your Empirical Project on Canvas. Your submission should be a single PDF file containing three parts:  
1. A 4-6 page research summary (double spaced and including references, graphs, and tables)  
2. A copy of do-file with your STATA code or an .R script file with your R code  
3. A copy of the log file (or screen output) of your STATA or R output  

# Specific questions to address in your research summary  

# Research Analysis

```{r}
# loading the data into R
ohp_df = here("data", "ohp.dta") |> read.dta()
```


### 1. Explain the difference between the variables treatment and ohp_all_ever_survey. Explain why treatment is the treatment variable ($D_{i}$), rather than ohp_all_ever_survey.  

Treatment is $D_{i}$ rather than ohp_all_ever_survey because of the constrained medicaid resources available to the experiment as well as the restriction of the experiment to a specific time period. ohp_all_ever_survey documents the number of people that have been enrolled in medicaid throughout all of time in Oregon, which may not be the case during the course of the study.

### 2. Provide evidence that the OHP lottery really did randomly assign individuals to treatment and control groups. Similar to Table 1 in Taubman et al (2014), please create a nicely formatted table that reports means of 4 to 6 relevant characteristics for individuals in the control group.  

Part of this question is to get you to think about which variables should be balanced in a randomized experiment. You need to read carefully through all the variables in the dataset documentation attached at the end of this file and decide which 4 to 6 you will summarize.  

```{r}
# specifying 6 different models
b_1 = lm(age_inp ~ factor(treatment), data = ohp_df)
# do I need to cluster in order to find accurate standard errors
  se_age = sqrt(diag(vcovHC(b_1, cluster = ohp_df$cluster)))
                
b_2 = lm(chl_inp ~ factor(treatment), data = ohp_df)
  se_chl = sqrt(diag(vcovHC(b_2, cluster = ohp_df$cluster)))
  
b_3 = lm(doc_num_mod_inp ~ factor(treatment), data = ohp_df)
  se_doc = sqrt(diag(vcovHC(b_4, cluster = ohp_df$cluster)))
  
b_4 = lm(bp_sar_inp ~ factor(treatment), data = ohp_df)
  se_sar = sqrt(diag(vcovHC(b_5, cluster = ohp_df$cluster)))
  
b_5 = lm(rx_num_mod_inp ~ factor(treatment), data = ohp_df)
  se_rx = sqrt(diag(vcovHC(b_6, cluster = ohp_df$cluster)))
```


```{r, asis}
# forming a stargazer table for each of the balanced treatments
stargazer::stargazer(b_1,b_2,b_3,b_4,b_5,
          se = list(se_age,se_chl,se_doc,se_sar, se_rx),
          keep.stat=c("n","rsq"),
          model.numbers = FALSE,
          dep.var.labels=c("Age","","2006 Primary","Age"), 
          type = 'text', style = "qje",
          covariate.labels = c("Treatment"))
```

### 3. For each of the variables you summarized above, calculate:  
i. the difference between the mean in the treatment group and the mean in the control group  
ii. the standard error for the difference in means.  
Add these as columns two and three to the table you started in question 2.  

### 4. Is the balance table consistent with individuals having been randomly assigned to treatment group and control groups? Why or why not?  

### 5. Estimate the compliance rate for the OHP experiment. That is, what is the effect of being assigned to the treatment group on the probability of being enrolled in Medicaid?  

Hint: For this question and question 7, you can use the same regression as in question 3, just changing the dependent variable. 

### 6. What is the intent-to-treat (ITT) effect of the OHP experiment on health outcomes? Please create a nicely formatted table that reports ITT estimates on 4 to 6 relevant health outcomes. Again, part of this question is to get you to think about which 4 to 6 variables could be used as health outcome variables.  

### 7. What is the “treatment on the treated” effect (ATET) of the OHP experiment, i.e. the effect among those who applied for Medicaid? Estimate it for every health outcome you chose in question 6 and provide some intuition for the calculation of this estimate.  

### 8. Do you have to worry about attrition bias in analyzing this data? Explain why or why not.  

### 9. Suppose that you are submitting these results to a general interest journal such as Science for publication. Write an abstract of 200 or fewer words describing what you have found in your analysis of the OHP data, similar to the abstract in Taubman et al. (2014).  

# Research Summary




