---
title: "lecture_notes"
author: "Nathan Bracken"
date: "3/29/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(here, data.table)
```

# EC 425 Notes

# Hal Varian, chief economist at google 
- We want to take data and make it more interpretable with regards to the world.

# The economist says that the most valuable resource in the world is data 
- We are interested in understanding if we want to spend money to increase education
- We want our spending to translate to some sort of outcome 
- depending on who you work for, you will have different priorities between causal inference and prediction

# Thinking about comparisons
- Units will be a priority for udnerstanding
- The difficult part is finding people or controls that are similar to the others
  - Randomizing the assignment of treatment

# Key Skills
- Discern between correllation and causation
- Identify which problems require prediction and which problems require causal inference
  - Critical thinking
  - Methods
  - Communication

Orient towards applied practitioners, emphasize on research design

# Why Regression?
All models are wrong, but some are useful.

# All modeling exercizes can be broadly categorized into one of two classes. 
1. Regression, continous outcome
  - conditional expectation of Y given X.
  - we are estimating the conoditional expectation function

# Why the conditionoal expectation function
  - The $E[\epsilon | X] = 0$ 
  -  The $E[Y | X]$ minimizes the mean squared error
     - function of X, 

# We want to find a line that minimizes the errors to the averages
    - orthog projections

# Why do we just take the conditional mean
   - who is this average person? why do we focus on conditional mean? 
    - we are focusing heavily on the lienar approximation of the conditional mean function
    - heavily penalizes the observations of outliers     
  - it is a useful baseline case
    - roger Koenker - every parameter woould like to grow up to be a distribution
    
# In class example 
```{r}
fm_df = here("data", "cleaned_Freddie_Mac.Rdata") |> load()
```
- code is available in the slides




